A Logical Solution of the Linearity Challenge

### 2025-08-01 paper/DeadlockFreedom.lagda.tex
drafted deadlock freedom

\begin{code}
open import Data.Sum
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.Bool using (Bool; true; false; if_then_else_)
open import Data.List.Base using ([]; _∷_; [_])
open import Relation.Nullary using (¬_; contradiction)
open import Relation.Binary.PropositionalEquality using (refl)

open import Type
open import Context
open import Process
open import Reduction
open import Congruence
open import Permutations

data Cut {Γ} : Process Γ → Set where
  cut :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Cut (cut d p P Q)

data Link {Γ} : Process Γ → Set where
  link :
    ∀{A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Link (link d p)

data Input : ∀{Γ} → Process Γ → Set where
  fail :
    ∀{Γ Δ}
    (p : Γ ≃ [] + Δ) → Input (fail (split-l p))
  wait :
    ∀{Γ Δ} (p : Γ ≃ [] + Δ) {P : Process Δ} → Input (wait (split-l p) P)
  case :
    ∀{Γ Δ A B} (p : Γ ≃ [] + Δ) {P : Process (A ∷ Δ)} {Q : Process (B ∷ Δ)} →
    Input (case (split-l p) P Q)
  join :
    ∀{Γ Δ A B} (p : Γ ≃ [] + Δ) {P : Process (B ∷ A ∷ Δ)} →
    Input (join (split-l p) P)

data Client : ∀{Γ} → Process Γ → Set where
  client :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
    {P : Process (A ∷ Δ)} →
    Client (client (split-l p) P)
  weaken :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
    {P : Process Δ} →
    Client (weaken {A = A} (split-l p) P)
  contract :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
    {P : Process (¿ A ∷ ¿ A ∷ Δ)} →
    Client (contract (split-l p) P)

data Output : ∀{Γ} → Process Γ → Set where
  close : Output close
  select :
    ∀{Γ Δ A B} (x : Bool) (p : Γ ≃ [] + Δ) {P : Process ((if x then A else B) ∷ Δ)} →
    Output (select x (split-l p) P)
  fork :
    ∀{Γ Δ Δ₁ Δ₂ A B} (p : Γ ≃ [] + Δ) (q : Δ ≃ Δ₁ + Δ₂)
    {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
    Output (fork (split-l p) q P Q)

data Delayed : ∀{Γ} → Process Γ → Set where
  fail :
    ∀{A Γ Δ}
    (p : Γ ≃ ⊤ , Δ) → Delayed (fail (split-r {A} p))
  wait :
    ∀{C Γ Δ} (p : Γ ≃ ⊥ , Δ) {P : Process (C ∷ Δ)} → Delayed (wait (split-r p) P)
  case :
    ∀{Γ Δ C A B} (p : Γ ≃ A & B , Δ) {P : Process (A ∷ C ∷ Δ)} {Q : Process (B ∷ C ∷ Δ)} →
    Delayed (case (split-r p) P Q)
  join :
    ∀{Γ Δ C A B} (p : Γ ≃ A ⅋ B , Δ) {P : Process (B ∷ A ∷ C ∷ Δ)} →
    Delayed (join (split-r p) P)
  select :
    ∀{Γ Δ C A B} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) {P : Process ((if x then A else B) ∷ C ∷ Δ)} →
    Delayed (select x (split-r p) P)
  fork-l :
    ∀{Γ Δ Δ₁ Δ₂ C A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Δ₁ + Δ₂)
    {P : Process (A ∷ C ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
    Delayed (fork (split-r p) (split-l q) P Q)
  fork-r :
    ∀{Γ Δ Δ₁ Δ₂ C A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Δ₁ + Δ₂)
    {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ C ∷ Δ₂)} →
    Delayed (fork (split-r p) (split-r q) P Q)
  client :
    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (A ∷ C ∷ Δ)} →
    Delayed (client (split-r p) P)
  weaken :
    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (C ∷ Δ)} →
    Delayed (weaken (split-r p) P)
  contract :
    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (¿ A ∷ ¿ A ∷ C ∷ Δ)} →
    Delayed (contract (split-r p) P)

data Server : ∀{Γ} → Process Γ → Set where
  server :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ) (un : Un Δ) {P : Process (A ∷ Δ)} →
    Server (server (split-l p) un P)

data DelayedServer : ∀{Γ} → Process Γ → Set where
  server :
    ∀{Γ Δ A C} (p : Γ ≃ ¡ A , Δ) (un : Un Δ) {P : Process (A ∷ ¿ C ∷ Δ)} →
    DelayedServer (server (split-r p) (un-∷ un) P)

Thread : ∀{Γ} → Process Γ → Set
Thread P = Link P ⊎ Delayed P ⊎ Output P ⊎ Input P ⊎ Client P ⊎ Server P ⊎ DelayedServer P

link-thread : ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Thread (link d p)
link-thread d p = inj₁ (link d p)

@@

server-thread :
  ∀{Γ Δ A} (p : Γ ≃ [ ¡ A ] + Δ) (un : Un Δ) {P : Process (A ∷ Δ)} →
  Thread (server p un P)
server-thread (split-l p) un = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (server p un))))))
server-thread (split-r p) (un-∷ un) = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (server p un))))))

data CanonicalCut {Γ} : Process Γ → Set where
  cc-link :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Link P → CanonicalCut (cut d p P Q)
  cc-delayed :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Delayed P → CanonicalCut (cut d p P Q)
  cc-delayed-server :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    DelayedServer P → Server Q → CanonicalCut (cut d p P Q)
  cc-redex :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Output P → Input Q → CanonicalCut (cut d p P Q)
  cc-connect :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Server P → Client Q → CanonicalCut (cut d p P Q)

output-output :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × Output Q)
output-output d-𝟙-⊥ (close , ())

@@

delayed-server-delayed-served :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (DelayedServer P × DelayedServer Q)
delayed-server-delayed-served () (server p un , server p₁ un₁)

canonical-cut :
  ∀{Γ Γ₁ Γ₂ A B}
  {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)}
  (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
  Thread P → Thread Q →
  ∃[ R ] CanonicalCut R × cut d p P Q ⊒ R
canonical-cut dc pc (inj₁ x) Qt = _ , cc-link dc pc x , s-refl
canonical-cut dc pc (inj₂ x) (inj₁ y) = _ , cc-link (dual-symm dc) (+-comm pc) y , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₁ x)) (inj₂ y) = _ , cc-delayed dc pc x , s-refl
canonical-cut dc pc (inj₂ (inj₂ x)) (inj₂ (inj₁ y)) = _ , cc-delayed (dual-symm dc) (+-comm pc) y , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₁ y))) = contradiction (x , y) (output-output dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = _ , cc-redex dc pc x y , s-refl
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (output-client dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (output-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (output-delayed-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₁ y))) = _ , cc-redex (dual-symm dc) (+-comm pc) y x , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-client (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-delayed-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (x , y) (input-input dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (input-client dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (input-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (input-delayed-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-client (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-delayed-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (client-client dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-connect (dual-symm dc) (+-comm pc) y x , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (client-delayed-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = _ , cc-connect dc pc x y , s-refl
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (y , x) (client-delayed-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (server-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = _ , cc-delayed-server (dual-symm dc) (+-comm pc) y x , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-delayed-server dc pc x y , s-refl
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (delayed-server-delayed-served dc)

Observable : ∀{Γ} → Process Γ → Set
Observable P = ∃[ Q ] P ⊒ Q × Thread Q

Live : ∀{Γ} → Process Γ → Set
Live P = Observable P ⊎ Reducible P

DeadlockFree : ∀{Γ} → Process Γ → Set
DeadlockFree {Γ} P = ∀(Q : Process Γ) → P ⇒ Q → Live Q

⊒Live : ∀{Γ} {P Q : Process Γ} → P ⊒ Q → Live Q → Live P
⊒Live pcong (inj₁ (_ , x , th)) = inj₁ (_ , s-tran pcong x , th)
⊒Live pcong (inj₂ (_ , red)) = inj₂ (_ , r-cong pcong red)

>>>>

\subsection{Deadlock Freedom}
\label{sec:deadlock-freedom-agda}

As we have seen in \Cref{sec:calculus}, the deadlock freedom result
(\Cref{thm:df}) rests on some notions and predicates about processes which we
now formalize in Agda. First of all we need to distinguish \emph{threads} from
\emph{cuts}. As it turns out, in the formalization it is convenient to provide
an even more fine-grained classification of threads, distinguishing between
links, input/output actions on linear/shared channels, and sometimes also
whether such actions operate on free or bound channels. We define predicates for
each of these process classes:

\begin{code}[hide]
open import Data.Sum
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.Bool using (Bool; true; false; if_then_else_)
open import Data.List.Base using ([]; _∷_; [_])
open import Relation.Nullary using (¬_; contradiction)
open import Relation.Binary.PropositionalEquality using (refl)

open import Type
open import Context
open import Permutations
open import Process
open import Reduction
open import Congruence
\end{code}

\begin{code}
data Cut            : ∀{Γ} → Process Γ → Set
data Link           : ∀{Γ} → Process Γ → Set
data Input          : ∀{Γ} → Process Γ → Set
data Output         : ∀{Γ} → Process Γ → Set
data Delayed        : ∀{Γ} → Process Γ → Set
data Client         : ∀{Γ} → Process Γ → Set
data Server         : ∀{Γ} → Process Γ → Set
data DelayedServer  : ∀{Γ} → Process Γ → Set
\end{code}

These predicates are relatively uninteresting since they are essentially
isomorphic to the relevant parts of the \AgdaDatatype{Process} datatype. The
only remarkable aspect is that \AgdaDatatype{Input}, \AgdaDatatype{Output},
\AgdaDatatype{Client} and \AgdaDatatype{Server} the channel being acted upon by
the thread is the \emph{first} in $\Context$, hence it is the \emph{most
recently introduced} channel, whereas in \AgdaDatatype{Delayed} and
\AgdaDatatype{DelayedServer} the channel is older. This allows us to distinguish
those threads that, in the context of a cut, operate on the channel bound by the
cut or on a free channel. As an illustration, consider the
\AgdaInductiveConstructor{wait} case in \AgdaDatatype{Input} and in
\AgdaDatatype{Delayed}. In the former predicate we have

\begin{AgdaAlign}
\begin{code}[hide]
data Cut where
  cut :
    ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Cut (cut d p P Q)

data Link where
  link :
    ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Link (link d p)

data Input where
  fail :
    ∀{Γ Δ}
    (p : Γ ≃ [] + Δ) → Input (fail (split-l p))
\end{code}

\begin{code}
  wait : ∀{Γ Δ} (p : Γ ≃ [] + Δ) {P : Process Δ} → Input (wait (split-l p) P)
\end{code}
where the \AgdaInductiveConstructor{split-l} constructor indicates that the
thread operates on the most recent channel. In the latter predicate we have

\begin{code}[hide]
  case :
    ∀{Γ Δ A B} (p : Γ ≃ [] + Δ) {P : Process (A ∷ Δ)} {Q : Process (B ∷ Δ)} →
    Input (case (split-l p) P Q)
  join :
    ∀{Γ Δ A B} (p : Γ ≃ [] + Δ) {P : Process (B ∷ A ∷ Δ)} →
    Input (join (split-l p) P)

data Client where
  client :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
    {P : Process (A ∷ Δ)} →
    Client (client (split-l p) P)
  weaken :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
    {P : Process Δ} →
    Client (weaken {A = A} (split-l p) P)
  contract :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
    {P : Process (¿ A ∷ ¿ A ∷ Δ)} →
    Client (contract (split-l p) P)

data Output where
  close : Output close
  select :
    ∀{Γ Δ A B} (x : Bool) (p : Γ ≃ [] + Δ) {P : Process ((if x then A else B) ∷ Δ)} →
    Output (select x (split-l p) P)
  fork :
    ∀{Γ Δ Δ₁ Δ₂ A B} (p : Γ ≃ [] + Δ) (q : Δ ≃ Δ₁ + Δ₂)
    {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
    Output (fork (split-l p) q P Q)

data Delayed where
  fail :
    ∀{A Γ Δ}
    (p : Γ ≃ ⊤ , Δ) → Delayed (fail (split-r {A} p))
\end{code}

\begin{code}
  wait : ∀{C Γ Δ} (p : Γ ≃ ⊥ , Δ) {P : Process (C ∷ Δ)} →
         Delayed (wait (split-r p) P)
\end{code}
where the \AgdaInductiveConstructor{split-r} constructor indicates that the
thread operates on a channel other than the most recent one.

\begin{code}[hide]
  case :
    ∀{Γ Δ C A B} (p : Γ ≃ A & B , Δ) {P : Process (A ∷ C ∷ Δ)} {Q : Process (B ∷ C ∷ Δ)} →
    Delayed (case (split-r p) P Q)
  join :
    ∀{Γ Δ C A B} (p : Γ ≃ A ⅋ B , Δ) {P : Process (B ∷ A ∷ C ∷ Δ)} →
    Delayed (join (split-r p) P)
  select :
    ∀{Γ Δ C A B} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) {P : Process ((if x then A else B) ∷ C ∷ Δ)} →
    Delayed (select x (split-r p) P)
  fork-l :
    ∀{Γ Δ Δ₁ Δ₂ C A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Δ₁ + Δ₂)
    {P : Process (A ∷ C ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
    Delayed (fork (split-r p) (split-l q) P Q)
  fork-r :
    ∀{Γ Δ Δ₁ Δ₂ C A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Δ₁ + Δ₂)
    {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ C ∷ Δ₂)} →
    Delayed (fork (split-r p) (split-r q) P Q)
  client :
    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (A ∷ C ∷ Δ)} →
    Delayed (client (split-r p) P)
  weaken :
    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (C ∷ Δ)} →
    Delayed (weaken (split-r p) P)
  contract :
    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (¿ A ∷ ¿ A ∷ C ∷ Δ)} →
    Delayed (contract (split-r p) P)

data Server where
  server :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ) (un : Un Δ) {P : Process (A ∷ Δ)} →
    Server (server (split-l p) un P)

data DelayedServer where
  server :
    ∀{Γ Δ A C} (p : Γ ≃ ¡ A , Δ) (un : Un Δ) {P : Process (A ∷ ¿ C ∷ Δ)} →
    DelayedServer (server (split-r p) (un-∷ un) P)
\end{code}
\end{AgdaAlign}

Now the \AgdaDatatype{Thread} predicate is simply the disjunction of all the
previous ones.

\begin{code}
Thread : ∀{Γ} → Process Γ → Set
Thread P = Link P ⊎ Delayed P ⊎ Output P ⊎ Input P ⊎
           Client P ⊎ Server P ⊎ DelayedServer P
\end{code}

We can now define observability, reducibility and aliveness in the expected way:

\begin{code}
Observable : ∀{Γ} → Process Γ → Set
Observable P = ∃[ Q ] P ⊒ Q × Thread Q

Reducible : ∀{Γ} → Process Γ → Set
Reducible P = ∃[ Q ] P ↝ Q

Alive : ∀{Γ} → Process Γ → Set
Alive P = Observable P ⊎ Reducible P
\end{code}

\begin{code}[hide]
link-thread : ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Thread (link d p)
link-thread d p = inj₁ (link d p)

@@

server-thread :
  ∀{Γ Δ A} (p : Γ ≃ [ ¡ A ] + Δ) (un : Un Δ) {P : Process (A ∷ Δ)} →
  Thread (server p un P)
server-thread (split-l p) un = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (server p un))))))
server-thread (split-r p) (un-∷ un) = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (server p un))))))
\end{code}

In order to prove that every (well-typed) process is alive, it is convenient to
also define a ``canonical'' form for cuts, which is a form that matches at least
one of the left-hand-sides of the rules for structural pre-congruence or
reduction (\Cref{tab:semantics}). This is the notion where the fine-grained
classification of threads introduced earlier comes into play.

\begin{code}
data CanonicalCut {Γ} : Process Γ → Set where
  cc-link :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Link P → CanonicalCut (cut d p P Q)
  cc-delayed :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Delayed P → CanonicalCut (cut d p P Q)
  cc-delayed-server :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    DelayedServer P → Server Q → CanonicalCut (cut d p P Q)
  cc-redex :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Output P → Input Q → CanonicalCut (cut d p P Q)
  cc-connect :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Server P → Client Q → CanonicalCut (cut d p P Q)
\end{code}

A \emph{canonical cut} $\Cut\x{P}{Q}$ has one of these forms:
\begin{itemize}
\item $P$ is a link (\AgdaInductiveConstructor{cc-link}), or
\item $P$ operates on a channel other than $x$ and is not a server
(\AgdaInductiveConstructor{cc-delayed}), or
\item $P$ and $Q$ are servers and $P$ operates on a channel other than $x$
(\AgdaInductiveConstructor{cc-delayed-server}), or
\item $P$ performs an output on $x$ and $Q$ performs an input on $x$
(\AgdaInductiveConstructor{cc-redex} and \AgdaInductiveConstructor{cc-connect}).
\end{itemize}

\begin{code}[hide]
output-output :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × Output Q)
output-output d-𝟙-⊥ (close , ())

@@

delayed-server-delayed-served :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (DelayedServer P × DelayedServer Q)
delayed-server-delayed-served () (server p un , server p₁ un₁)
\end{code}

Every cut $\Cut\x{P}{Q}$ where both $P$ and $Q$ are threads can be rewritten
into a canonical cut using structural pre-congruence:

\begin{code}
canonical-cut :
  ∀{Γ Γ₁ Γ₂ A B} {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)}
  (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
  Thread P → Thread Q → ∃[ R ] CanonicalCut R × cut d p P Q ⊒ R
\end{code}

\begin{code}[hide]
canonical-cut dc pc (inj₁ x) Qt = _ , cc-link dc pc x , s-refl
canonical-cut dc pc (inj₂ x) (inj₁ y) = _ , cc-link (dual-symm dc) (+-comm pc) y , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₁ x)) (inj₂ y) = _ , cc-delayed dc pc x , s-refl
canonical-cut dc pc (inj₂ (inj₂ x)) (inj₂ (inj₁ y)) = _ , cc-delayed (dual-symm dc) (+-comm pc) y , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₁ y))) = contradiction (x , y) (output-output dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = _ , cc-redex dc pc x y , s-refl
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (output-client dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (output-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (output-delayed-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₁ y))) = _ , cc-redex (dual-symm dc) (+-comm pc) y x , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-client (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-delayed-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (x , y) (input-input dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (input-client dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (input-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (input-delayed-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-client (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-delayed-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (client-client dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-connect (dual-symm dc) (+-comm pc) y x , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (client-delayed-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = _ , cc-connect dc pc x y , s-refl
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (y , x) (client-delayed-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (server-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = _ , cc-delayed-server (dual-symm dc) (+-comm pc) y x , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-delayed-server dc pc x y , s-refl
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (delayed-server-delayed-served dc)

⊒Alive : ∀{Γ} {P Q : Process Γ} → P ⊒ Q → Alive Q → Alive P
⊒Alive pcong (inj₁ (_ , x , th)) = inj₁ (_ , s-tran pcong x , th)
⊒Alive pcong (inj₂ (_ , red)) = inj₂ (_ , r-cong pcong red)
\end{code}

It is easy to prove that every canonical cut is alive, either reducing it or
applying structural pre-congruence to rewrite it into a thread.
